Analysis of cloud enhancement events in a 30-year record of global solar irradiance at Thessaloniki, Greece

### 2024-09-08 article/article.tex
Commit 2024-09-08 19:10

We approximated the expected cloud-free \(\text{GHI}_\text{ref}\) with
the radiative transfer model uvspec, part of libRadtran
\citep{Emde2016}, similarly to the approach used by
\citet{Vamvakas2020}. In uvspec we used the solar spectrum of
\citet{Kurucz1994} in the range \(280\) to \(2500\,\text{nm}\), the
radiative transfer solver ``disort'' in ``pseudospherical'' geometry and
the ``LOWTRAN'' gas parameterization. The model was run for a range of
variables to create a look-up table (LUT) for the estimation of the
cloud-free reference irradiance for each individual observation of our
dataset. In this context, the model was run for SZAs in the range \(10\)
-- \(90^\circ\) with a step of \(0.2^\circ\) and for the atmospheric
profiles of the Air Force Geophysics Laboratory \citep{Anderson1986}
midlatitude summer and midlatitude winter, representative of the warm
and cold seasons.

@@

The cloud-free reference values discussed above are based on the
climatological AOD and WC; hence they cannot describe accurately the
long-term variation of \(\text{GHI}_\text{ref}\) due to long-term
changes in the two atmospheric constituents, mainly AOD. As reported by
\citet{Natsis2023}, there is a long-term brightening effect in the GHI
data of Thessaloniki for the period 1993 -- 2023, which for cloud-free
data was attributed to long-term changes in aerosol effects. Therefore,
an adjustment of the \(\text{GHI}_\text{ref}\) during the period of
study was made using RTM simulations to account for the long-term
variations of the AOD. High quality AOD data with a Cimel sun-photometer
(part of AERONET) at Thessaloniki start only in 2003, while spectral AOD
measurements using direct-sun observations with a MKIII Brewer
spectrophotometer are available for the period 1997 -- 2017. These data
are taken sporadically during each day and are less dense compared to
the AERONET data. By comparing monthly AOD data at \(340\,\text{nm}\) of
the two instruments for the common periods of operation we adjusted the
Brewer data and filled the missing months of the Cimel record data with
Brewer data. Using the monthly time series of AOD at \(340\,\text{nm}\),
as well as monthly climatological values of the Angstrom exponent and
constant WC of \(15.6\,\text{mm}\) derived from the Cimel record, we
simulated with the RTM the cloud-free GHI at SZA of \(55^\circ\) for
each month in the period 1997 -- 2023. The SZA of \(55^\circ\) was
chosen as representative of all days in the year to get a rough estimate
of the annually averaged change in cloud-free GHI. A second-degree
polynomial fit was applied to the simulated yearly averaged GHI to
derive the long-term change in GHI due to aerosols
(Equation\nobreakspace{}\ref{eq:AODchange}).

As AERONET data start only in 2003, we used for the period 1994 -- 2005
estimates of changes in AOD at \(340\,\text{nm}\) derived from a
collocated Brewer spectrophotometer \citep{Kazadzis2007} to calculate
the trend in \(\text{GHI}_\text{ref}\) due to aerosols during this
period.

According to \citet{Kazadzis2007}, in the period 1997 -- 2005 the mean
AOD at \(340\,\text{nm}\) is \(0.4\) with a change of
\(-3.8\pm0.93\,\%\) per year, corresponding to a change of \(0.0153\)
per year. Using an Ångström coefficient \(\alpha = 1.6\), this
translates to a change in the Ångström coefficient \(\beta\) of
\(0.00272\) per year (or \(\beta=0.084\) in 1997 and \(\beta=0.059\) in
2005). Simulations with uvspec for the above Ångström coefficients,
assuming constant WC of \(15.6\,\text{mm}\) taken from the Cimel, and
for a SZA of \(55^\circ\) reveal a change of \(+0.21\,\%\) per year in
\(\text{GHI}_\text{ref}\). The SZA of \(55^\circ\) was chosen as
representative of all days in the year in order to get a rough estimate
of the annually averaged change in cloud-free irradiance. For the period
2005 -- 2023 we used the mean monthly values of AOD and WC from AERONET
in a similar simulation scheme to calculate the monthly mean cloud-free
irradiance, and finally the change of \(+0.14\,\%\) per year. We applied
these two long-term changes (see
Figure\nobreakspace{}\ref{fig:CS-change}) to the climatological
\(\text{GHI}_\text{ref}\), to create a more realistic representation of
the cloud-free irradiance for the whole period of study. The rate of
changes in the two subperiods are different, but reasonable, since the
effect of the enforced air pollution abatement measures in the area in
the 1990s had greater effects in the first years of application.

\begin{equation}
\Delta(\text{AOD}) [\%] = -12170 + 12.05 \cdot y + -0.002981 \cdot y^2 \label{eq:AODchange}
\end{equation} where \(y\) the date as a decimal fraction of the year.
For the period of the study the relative change of AOD ranges between
\(-1.88\) and \(-0.23\,\%\).

>>>>

We approximated the expected cloud-free \(\text{GHI}_\text{ref}\) with
the radiative transfer model uvspec, part of libRadtran
\citep{Emde2016}, similarly to the approach used by
\citet{Vamvakas2020}. In uvspec we used the solar spectrum of
\citet{Kurucz1994} in the range \(280\) to \(2500\,\text{nm}\), the
radiative transfer solver ``disort'' in ``pseudospherical'' geometry and
the ``LOWTRAN'' gas parameterization. The model was run for
climatological values of the Angstrom coefficients, water column (WC),
SZA and the appropriate seasonal atmospheric profile to create a look-up
table (LUT) for the estimation of the cloud-free reference irradiance
for each individual observation of our dataset. In this context, the
model was run for SZAs in the range \(10\) -- \(90^\circ\) with a step
of \(0.2^\circ\) and for the atmospheric profiles of the Air Force
Geophysics Laboratory \citep{Anderson1986} midlatitude summer and
midlatitude winter, representative of the warm and cold seasons.

@@

The cloud-free reference values discussed above are based on the
climatological AOD and WC; hence they cannot describe accurately the
long-term variation of \(\text{GHI}_\text{ref}\) due to long-term
changes in the two atmospheric constituents, mainly AOD. As reported by
\citet{Natsis2023}, there is a long-term brightening effect in the GHI
data of Thessaloniki for the period 1993 -- 2023, which for cloud-free
data was attributed to long-term changes in aerosol effects. Therefore,
an adjustment of the \(\text{GHI}_\text{ref}\) during the period of
study was made using RTM simulations to account for the long-term
variations of the AOD. High quality AOD data with a Cimel sun-photometer
(part of AERONET) at Thessaloniki start only in 2003, while spectral AOD
measurements using direct-sun observations with a MKIII Brewer
spectrophotometer are available for the period 1997 -- 2017. These data
are taken sporadically during each day and are less dense compared to
the AERONET data. By comparing monthly AOD data at \(340\,\text{nm}\) of
the two instruments for the common periods of operation we adjusted the
Brewer data and filled the missing months of the Cimel record data with
Brewer data. Using the monthly time series of AOD at \(340\,\text{nm}\),
as well as monthly climatological values of the Ångström exponent and
constant WC of \(15.6\,\text{mm}\) derived from the Cimel record, we
simulated with the RTM the cloud-free GHI at SZA of \(55^\circ\) for
each month in the period 1997 -- 2023. The SZA of \(55^\circ\) was
chosen as representative of all days in the year to get a rough estimate
of the annually averaged change in cloud-free GHI. A second-degree
polynomial fit was applied to the simulated yearly averaged GHI to
derive the long-term change in GHI due to aerosols: \begin{equation}
\Delta(\text{AOD}) [\%] = -12170 + 12.05 \cdot y + -0.002981 \cdot y^2 \label{eq:AODchange}
\end{equation} where \(y\) is the date as a decimal fraction of the
year. For the period of the study the relative change of AOD ranges
between \(-1.88\) and \(-0.23\,\%\). Finally, we applied these long-term
changes of Equation\nobreakspace{}\ref{eq:AODchange} to the
climatological \(\text{GHI}_\text{ref}\), to create a more realistic
representation of the cloud-free irradiance for the entire period of
study. For the period 1994 -- 1996 where no AOD data are available, we
assumed that the changes in GHI follow the same polynomial fit.
